---
title: "Tutorial: Using XGBoost for Species-Distribution-Modelling"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Dummy notes area

https://juliasilge.com/blog/baseball-racing/
XGBoost-Docs: https://xgboost.readthedocs.io/en/stable/
https://dieghernan.github.io/tidyterra/reference/geom_spatraster.html
https://www.youtube.com/watch?v=gKyUucJwD8U&list=WL&index=71



# Species Distribution Modeling




# Extreme Gradient Boosting Algorithem




# Application of XGBoost




## Prerequisites

```{r Install packages, eval=FALSE}
install.packages('tidyverse')
install.packages("ggplot2")
install.packages('tidyterra')
install.packages('fastDummies')
install.packages('tidymodels')

#if(!require(dplyr)){install.packages('dplyr')}
```

Install XGBoost

```{r Setup from Cran, eval=FALSE, include=TRUE}
install.packages("xgboost")
```
To use GPU acceleration you need to install a prebuild binary:
```{r Manuall setup for GPU acceleration, eval=FALSE, include=TRUE}
# Check if r is in path
system("where R")
# TODO: Add r to session path here if not found else you have to edit your user profile path settings in windows and add r manually
# Install dependencies
system('R -q -e "install.packages(c(\'data.table\', \'jsonlite\'))"')
# Install XGBoost
system(paste("R CMD INSTALL ", getwd(),  "/xgboost_r_gpu_win64_21d95f3d8f23873a76f8afaad0fee5fa3e00eafe.tar.gz", sep=""))
```


```{r loading libaries}
require(xgboost)
require(terra)
require(geodata)

require(dplyr)
require(tidyverse)
require(ggplot2)
require(tidyterra)
require(fastDummies)
require(tidymodels)
require(vip)
```

## Preparing data

Species Occurrences data from ...

```{r perparing species data}

species_occurrences_all = read.table("data/PakistanLadakh.csv", sep=",", header=TRUE)
species_occurrences_all = sf::st_as_sf(species_occurrences_all, coords=c("x", "y"), remove=TRUE, crs=sf::st_crs("epsg:4326"))

# list of species 
species_names = (species_occurrences_all %>% distinct(species) %>% select(species))$species

# list of species ordered by number of samples:
species_nsamples = data.frame(species_occurrences_all) %>% count(species, sort=TRUE) %>% rename(n_samples = n)
```

Environmental data used for modelling

Using the package geodata to obtain environmental data of pakistan. The ISO country code for pakistan can be found by using geodata::country_codes(query = 'Pakistan')

Show List of bioclim data variables

```{r preparing environmental data}
border <- geodata::gadm(country='PAK', level=0, path='./data')


worldclim_bio = geodata::worldclim_country(country="PAK", res=10, var="bio", path="data/", version = "2.1")
# TODO: rename layers to more meaningful names 
names(worldclim_bio)<-substr(names(worldclim_bio), 11, 20)

# mask to border of Pakistan
worldclim_bio = terra::mask(worldclim_bio, border)

elevation = geodata::elevation_30s(country='PAK', path='data/')
```

```{r Plotting species occurrences}
# select which species you want to plot by commenting in/out desired lines
#species_plot = species_occurrences_all
species_plot = dplyr::filter(species_occurrences_all, species=="Aglais_caschmirensis")
#species_Plot = dplyr::filter(species %in% species_names[1:10])

# TODO: finalize legend
ggplot() + 
  geom_spatraster(data = elevation) + 
  scale_fill_hypso_c() + 
  geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) + 
  geom_sf(data = species_plot, mapping = aes(colour = species) );
```

```{r Plotting grouped by species into single maps, eval=FALSE, include=FALSE, fig.width=50}
###                        ###
### not targeting the goal ###
###                        ###  
# Filtering for multiple species: 
# species_plot
#species_plot = species_occurrences_all %>% filter(species %in% species_names[1:10])

# INFO: genearting the plot takes a lot timee due to the high number of different species
ggplot(species_plot) + 
  geom_spatraster(data = elevation) + 
  scale_fill_hypso_c() + 
  geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) + 
  geom_sf(data = species_plot, mapping = aes(colour = species) ) +
  facet_wrap(facets = vars(species), ncol = 2);
```



Extract environmental data for presence and absence points

1. Extract presence points of observed occurrences with bioclim and elevation data
2. Generate background points with bioclim and elevatoin data with no obserations
3. Join presence and absence points together
4. Generate dummy columns for the species

# TODO: Write in Text: Problem of tiny distance between absence points and presence points

```{r}
# TODO: some sample points are outside of pakistan, remove them??

species_presence = species_occurrences_all #%>% filter(species %in% species_names[1:10])
#species_presence = sf::st_as_sf(terra::extract(worldclim_bio,species_occurrences, bind=TRUE))

# generate distributed background points with col 'occurrence' = 0
# TODO: why 1000 points?? give a explanation for the decisison
background_points = sf::st_sample(sf::st_as_sf(border), size = 1000)
# add species = NA to the backgroundpoints, needed for rbind to join the data
species_absence = cbind(background_points, data.frame(species = as.character(NA)))
species_absence = sf::st_as_sf(species_absence)

modeling_data = rbind(species_presence, species_absence)
# check if data is joined correctly, head should contain species names, tail NA values in the species column
head(modeling_data); tail(modeling_data)

# Extract values from worldclim_bio and elevation and join them to our modeling_data
extraction_worldclim_bio = terra::extract(worldclim_bio,modeling_data, bind=FALSE, ID=FALSE)
extraction_elevation = terra::extract(elevation,modeling_data, bind=FALSE, ID=FALSE)
modeling_data_extracted = cbind( modeling_data, extraction_worldclim_bio, extraction_elevation)

# Convert 
# TODO: maybe this can be done by the workflow
final_modeling_data <- dummy_columns( modeling_data_extracted, select_columns = 'species', ignore_na = FALSE )
final_modeling_data <- 
  final_modeling_data %>% 
  mutate_at(c(23,ncol(final_modeling_data)), ~tidyr::replace_na(.,0))
  #mutate(species_Aglais_caschmirensis = ifelse(is.na(species_Aglais_caschmirensis), 0, species_Aglais_caschmirensis)) %>%
  #mutate( species_Aglais_caschmirensis = factor(species_Aglais_caschmirensis))

head(final_modeling_data)
str(final_modeling_data)


#########################################################################################
###################################### Plots ############################################
#########################################################################################


# plot of final_modeling_data
ggplot() + 
  geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) + 
  geom_sf(data = sf::st_as_sf(final_modeling_data), mapping=aes(color=species), show.legend = FALSE)
```

```{r, eval=F, include=F}
#############  old ###################
#############   |  ###################
#############   V  ###################

species_occurrences=dplyr::filter(species_occurrences_all, species=="Aglais_caschmirensis")


species_presence = sf::st_as_sf(terra::extract(worldclim_bio,species_occurrences, bind=TRUE))
tmp = terra::extract(elevation, species_occurrences, bind=FALSE, ID=FALSE)
species_presence = cbind(species_presence, tmp)
head(species_presence)

# generate distributed background points with col 'occurrence' = 0
# TODO: why 1000 points?? give a explanation for the decisison
background_points = sf::st_sample(sf::st_as_sf(border), size = 1000)

ggplot() + 
  geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) + 
  geom_sf(data = background_points)

# TODO: more effiecently would be the bind presence and absence points together first and then extract values from raster and elevation
# Read values from bio
species_absence = sf::st_as_sf(terra::extract(worldclim_bio, terra::vect(background_points), bind=TRUE))
tmp = terra::extract(elevation, terra::vect(background_points), bind=FALSE, ID=FALSE)
tmp2 = data.frame(c(NA))
species_absence = cbind(species_absence, tmp, tmp2); rm(tmp); rm(tmp2)
species_absence <- species_absence %>% rename(species = 'c.NA.')
head(species_absence)

# cols have to be identical to rbind data frames together, check this before
str(species_presence)
str(species_absence)
modeling_data_raw = rbind(species_presence, species_absence)
# check if join was succesfull by head and tail of occurrence column
head(modeling_data_raw$occurrence); tail(modeling_data_raw$occurrence)

head(modeling_data_raw)

# TODO: convert species into logical presence factors
modeling_data= dummy_columns( modeling_data_raw, select_columns = 'species', ignore_na = TRUE )
modeling_data <- 
  modeling_data %>% 
  replace(is.na(starts_with(species_)), 0)
  #mutate(species_Aglais_caschmirensis = ifelse(is.na(species_Aglais_caschmirensis), 0, species_Aglais_caschmirensis)) %>%
  #mutate( species_Aglais_caschmirensis = factor(species_Aglais_caschmirensis))

head(modeling_data)
str(modeling_data)
modeling_data
```

Splitting the data into test and verfication dataset

## Data Overview

## Training the model

```{r}
# tracking runtime
start.time <- Sys.time()
# current runtime for all species ~35min

# use this for development to shorten runtime
#final_modeling_data_save = final_modeling_data
#final_modeling_data = final_modeling_data_save
#final_modeling_data = final_modeling_data %>% filter(species %in% species_names[1:50] | is.na(species))

for( sp in (final_modeling_data %>% distinct(species) %>% filter(!is.na(species)))$species )
{
  print(paste("Sarting compute for species:", sp))
  tryCatch(
    {
      # TODO: select only used cols
      data = final_modeling_data %>% 
        filter(species == sp | is.na(species)) %>% 
        select(ends_with(sp), matches("elv"), starts_with("bio")) %>%
        rename_at( 1, ~"species_occurrence") %>%
        mutate(species_occurrence = ifelse(is.na(species_occurrence), 0, species_occurrence)) %>%
        mutate(species_occurrence = factor(species_occurrence))
      
      #str(data)
      #message(data)
      
      # using "final_modeling_data" because of missing geometry in "data"
      ggplot() + 
        geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) + 
        geom_sf(data = sf::st_as_sf(final_modeling_data %>% filter(species == sp)), mapping=aes(color=species))
      # Save the plot for every species in out/<species>/<species>.png
      ggsave(paste(sp, ".png", sep=""), path=paste("out/",sp, sep=""))
      
      ########## Training the model ##########
      xgboost_recipe <- 
        recipe( species_occurrence ~ . , data = data)
      
      xgboost_model <- 
        boost_tree(
        trees = 500,
        min_n = 2,
        mtry = 3,
        learn_rate = 0.01
      ) %>%
        set_engine("xgboost") %>%
        set_mode("classification")

      xgboost_workflow <- workflow() %>% add_recipe(xgboost_recipe) %>% add_model(xgboost_model)

      xgboost_fit <- xgboost_workflow %>% 
        fit(data = data)
      
      xgboost_fit %>% 
        extract_fit_parsnip() %>%
        vip(geom = "point", num_features = 15)
      ggsave(paste(sp, "_fit.png", sep=""), path=paste("out/",sp, sep=""))
      
      
      # TODO: create new random point dataset of pakistan with environmental data and high density 
      # for a better spatial resultion of the prediction
      predicted = augment(xgboost_fit, final_modeling_data)
      
      predicted = sf::st_as_sf( predicted %>% select(!starts_with("species_")) )

      ggplot() + 
        geom_spatraster(data = elevation) + 
        scale_fill_hypso_c() + 
        geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) +
        #geom_sf(data = sf::st_as_sf(data), mapping = aes(color = 'bg')) + 
        geom_sf(data = predicted %>% filter(species == sp), mapping = aes(color = 'observed'), size = 5) +
        geom_sf(data = predicted %>% filter(.pred_class == 1), mapping = aes(color = 'predicted'), size = 3 )
      
      ggsave(paste(sp, "_model.png", sep=""), path=paste("out/",sp, sep=""))
      
    },
    error=function(cond)
    {
      message("Error:")
      message(cond)
    },
    warning=function(cond)
    {
      message("Warning:")
      message(cond)
    },
    finally={
      message("finally")
    }
  )
  print("finished")
}

# tracking runtime
end.time <- Sys.time()
diff <- end.time - start.time
diff

```

```{r, eval = F, include=F, fig.width = 15}

# only for development, filter to one species ## DO not filter here since absence points will be filtered out
modeling_data_save = modeling_data
modeling_data = modeling_data %>% select(c(-geometry, -species))
modeling_data
str(modeling_data)
# resulst in getting a dataset with 52 samples pf aglais caschirensis

predic_colname <- paste('species_', species_names[6], sep="")

# taken from https://www.tidymodels.org/start/recipes/
xgboost_recipe <- 
  recipe( species_Aglais_caschmirensis ~ . , data = modeling_data)

#summary(sdm_recipe)

#prep(sdm_recipe)

xgboost_model <- 
  boost_tree(
    trees = 1000,
    min_n = 2,
    mtry = 3,
    learn_rate = 0.01
  ) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgboost_workflow <- workflow() %>% add_recipe(xgboost_recipe) %>% add_model(xgboost_model)

xgboost_fit <- xgboost_workflow %>% 
  fit(data = modeling_data)


library(vip)
xgboost_fit %>% 
  extract_fit_parsnip() %>%
  vip(geom = "point", num_features = 15)


out = augment(xgboost_fit, modeling_data_save )

out

out = sf::st_as_sf(out)

ggplot() + 
  geom_spatraster(data = elevation) + 
  scale_fill_hypso_c() + 
  geom_sf(data = sf::st_as_sf(border), mapping=aes(alpha = 0), show.legend = FALSE) +
  geom_sf(data = sf::st_as_sf(modeling_data_save), mapping = aes(color = 'bg')) + 
  geom_sf(data = out %>% filter(species_Aglais_caschmirensis == 1), mapping = aes(color = 'observed'), size = 5) +
  geom_sf(data = out %>% filter(.pred_class == 1), mapping = aes(color = 'predicted'), size = 3 )



```

```{r, eval = F, include=F}
###             ###
### not working ###
###             ###  

modeling_data

dtrain <- 
  xgb.DMatrix(
    data = as.matrix(modeling_data %>% select(-species_Aglais_caschmirensis)), 
    label = as.matrix(modeling_data %>% select(species_Aglais_caschmirensis))
    )


xgb_model <- xgboost(data = dtrain, max.depth = 2, eta = 1, nthread = 2, nrounds = 200, objective = "binary:logistic", verbose = T)

xgb.plot.shap( data = as.matrix(modeling_data %>% select(-species_Aglais_caschmirensis)), model = xgb_model, top_n=7)

#xgboost_pred <- predict(model, as.matrix(species_absence))



```

## Model results

Species Richness Map


# Conclusion

How was Computing Speed compared to other methods? 


print sessioninfo