---
title: "Tutorial: Using XGBoost for Species-Distribution-Modelling"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Species Distribution Modeling


# Extreme Gradient Boosting Algorithem


# Application: XGBoost


## Prerequisites

Install XGBoost
XGBoost-Docs: https://xgboost.readthedocs.io/en/stable/

```{r Setup from Cran, eval=FALSE, include=TRUE}
install.packages("xgboost")
```

```{r Manuall setup for GPU acceleration, eval=FALSE, include=TRUE}
# Check if r is in path
system("where R")
# TODO: Add r to session path here if not found else you have to edit your user profile path settings in windows and add r manually
# Install dependencies
system('R -q -e "install.packages(c(\'data.table\', \'jsonlite\'))"')
# Install XGBoost
system(paste("R CMD INSTALL ", getwd(),  "/xgboost_r_gpu_win64_21d95f3d8f23873a76f8afaad0fee5fa3e00eafe.tar.gz", sep=""))
```


```{r}
require(xgboost)
require(terra)
require(geodata)
```

## Preparing data

Species Occurrences data from ...

```{r perparing species data}

species_occurrences = read.table("data/PakistanLadakh.csv", sep = ",", header = TRUE)
head(species_occurrences)
#str(species_occurrences)

species_occurences=dplyr::filter(species_occurrences, species=="Aglais_caschmirensis")
species_occurrences=sf::st_as_sf(species_occurrences, coords=c("x", "y"), remove=F, crs=sf::st_crs("epsg:4326"))
plot(species_occurrences)

```

Environmental data used for modelling

Using the package geodata to obtain environmental data of pakistan. The ISO country code for pakistan can be found by using geodata::country_codes(query = 'Pakistan')

```{r preparing environmental data}
border <- geodata::gadm(country='PAK', level=0, path='./data')

worldclim = geodata::worldclim_country(country="PAK", res=10, var="bio", path="data/", version = "2.1")
#names(worldclim)<-substr(names(environmental), 11, 20)
# mask to border of Pakistan
worldclim = terra::mask(worldclim, border)
#terra::writeRaster(r, "data/bioclim.tif", overwrite=T)

elevation = geodata::elevation_30s(country='PAK', path='data/')

plot(border)
plot(worldclim)
plot(elevation)
```

Extract environmental data for presence and absence points

```{r}
species_presence = terra::extract(worldclim,species_occurrences)
# species_absence = terra::extract(worldclim,background_points)

head(species_presence)
```

Splitting the data into test and verfication dataset


## Training the model

## Model results

